### 2025-03-13 paper/quarto_ecml/paper.tex
Review counterfactual training

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
\cite{altmeyer2023explaining}. In this context, domain constraints can
be imposed by simply projecting counterfactuals back to the specified
domain, if the previous gradient step resulted in updated feature values
that were out-of-domain. Mutability constraints can similarly be
enforced by setting partial derivatives to zero to ensure that features
are only mutated in the allowed direction, if at all.

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. Here it is
crucial that we avoid penalizing implausibility that arises due to
mutability constraints. For any mutability-constrained feature \(d\)
this can be achieved by enforcing
\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
violate mutability constraints. Specifically, we set
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Feature \(d\) is strictly immutable in practice.
\item
  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
  only be decreased in practice.
\item
  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
  only be increased in practice.
\end{enumerate}

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
(1) the subjects' \emph{age}, which we define as immutable, and (2) the
subjects' existing level of \emph{debt}, which we define as mutable.

We have simulated this scenario using synthetic data with independent
features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
outcomes for different training procedures using the same model
architecture each time (a linear classifier). In each case, we show the
linear decision boundary (green) and the training data colored according
to their ground-truth label: orange points belong to the target class,
\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
indicate counterfactuals in the target class generated at test time
using generic gradient descent until convergence.

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they are clearly
distinguishable from the ground-truth data. In panel (b), we have
trained our model with counterfactual training, once again not imposing
mutability constraints at test time. We observe that the counterfactuals
are clearly plausible, therefore meeting the first objective of
Definition~\ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals (very young applicants). By comparison,
counterfactual paths are shorter on average in panel (d), where we have
used counterfactual training and protected immutable features as
described in Section~\ref{sec-constraints}. In particular, we observe
that due to the classifier's lower sensitivity to \emph{age}, recourse
recommendations with respect to \emph{debt} are much more homogenous, in
that they do not disproportionately punish younger individuals. The
counterfactuals are also plausible with respect to the mutable feature.
Thus, we consider the model in panel (d) as the most explainable
according to Definition~\ref{def-explainability}.

@@

\caption{\label{fig-poc}Visual illustration of how counterfactual
training improves explainability. See Example~\ref{exm-grad} for
details.}

@@

Does our proposed counterfactual training objective
(Equation~\ref{eq-obj}) induce models to learn plausible explanations?

@@

Does our proposed counterfactual training objective
(Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
outcomes in the presence of actionability constraints?

@@

What are the effects of different hyperparameter choices with respect to
Equation~\ref{eq-obj}?

>>>>

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
\cite{altmeyer2023explaining}. In this context, domain constraints can
be imposed by simply projecting counterfactuals back to the specified
domain, if the previous gradient step resulted in updated feature values
that were out-of-domain. Mutability constraints can similarly be
enforced by setting partial derivatives to zero to ensure that features
are only perturbed in the allowed direction, if at all.

Since such actionability constraints are binding at test time, we should
also impose them when generating \(\mathbf{x}^\prime\) during each
training iteration to inform model representations. Through their effect
on \(\mathbf{x}^\prime\), both types of constraints influence model
outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
penalizing implausibility that arises due to mutability constraints. For
any mutability-constrained feature \(d\) this can be achieved by
enforcing \(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever
perturbing \(\mathbf{x}^\prime[d]\) in the direction of
\(\mathbf{x}[d]\) would violate mutability constraints. Specifically, we
set \(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Feature \(d\) is strictly immutable in practice.
\item
  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\), but feature \(d\) can
  only be decreased in practice.
\item
  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\), but feature \(d\) can
  only be increased in practice.
\end{enumerate}

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
features: (1) the subjects' \emph{age}, which we define as immutable,
and (2) the subjects' existing level of \emph{debt}, which we define as
mutable.

We have simulated this scenario using synthetic data with two
independent features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
outcomes for different training procedures using the same model
architecture each time (a linear classifier). In each case, we show the
decision boundary (in green) and the training data colored according to
their ground-truth label: orange points belong to the target class,
\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
indicate counterfactuals in the target class generated at test time
using generic gradient descent until convergence.

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they do not comply
with the distribution of the factual samples in the target class to the
point where they are clearly distinguishable from the ground-truth data.
In panel (b), we have trained our model with counterfactual training,
once again without any mutability constraints. We observe that the
counterfactuals are highly plausible, meeting the first objective of
Definition~\ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals (very young applicants). By comparison,
counterfactual paths are shorter on average in panel (d), where we have
used counterfactual training and protected the immutable feature as
described in Section~\ref{sec-constraints}. We observe that due to the
classifier's lower sensitivity to \emph{age}, recourse recommendations
with respect to \emph{debt} are much more homogenous and do not
disproportionately punish younger individuals. The counterfactuals are
also plausible with respect to the mutable feature. Thus, we consider
the model in panel (d) as the most explainable according to
Definition~\ref{def-explainability}.

@@

\caption{\label{fig-poc}Illustration for Example~\ref{exm-grad} of how
CT improves model explainability.}

@@

To what extent does our proposed counterfactual training objective
(Equation~\ref{eq-obj}) induce models to learn plausible explanations?

@@

To what extent does our proposed counterfactual training objective
(Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
outcomes in the presence of actionability constraints?

@@

What are the effects of hyperparameter selection with respect to
Equation~\ref{eq-obj}?

### 2025-03-13 paper/sections/experiments.qmd
final table done
Review counterfactual training

Does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?

@@

Does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?

@@

What are the effects of different hyperparameter choices with respect to @eq-obj?

>>>>

To what extent does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?

@@

To what extent does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?

@@

What are the effects of hyperparameter selection with respect to @eq-obj?

### 2025-03-13 paper/sections/method.qmd
Review counterfactual training

Many existing counterfactual explainers support domain and mutability constraints out-of-the-box. In fact, both types of constraints can be implemented for any counterfactual explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining]. In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain. Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only mutated in the allowed direction, if at all.

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if:

1. Feature $d$ is strictly immutable in practice.
2. We have $\mathbf{x}[d]>\mathbf{x}^\prime[d]$ but feature $d$ can only be decreased in practice.
3. We have $\mathbf{x}[d]<\mathbf{x}^\prime[d]$ but feature $d$ can only be increased in practice.

@@

It is worth highlighting that @prp-mtblty assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxy features that remain unprotected. We discuss this limitation in @sec-discussion.

## Illustration

@@

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.

We have simulated this scenario using synthetic data with independent features and Gaussian class-conditional densities in @fig-poc. The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier). In each case, we show the linear decision boundary (green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate counterfactuals in the target class generated at test time using generic gradient descent until convergence.

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with counterfactual training, once again not imposing mutability constraints at test time. We observe that the counterfactuals are clearly plausible, therefore meeting the first objective of @def-explainability.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints. In particular, we observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous, in that they do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.

@@

![Visual illustration of how counterfactual training improves explainability. See @exm-grad for details.](/paper/figures/poc.svg){#fig-poc}

>>>>

Many existing counterfactual explainers support domain and mutability constraints out-of-the-box. In fact, both types of constraints can be implemented for any counterfactual explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining]. In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain. Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.

Since such actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if:

1. Feature $d$ is strictly immutable in practice.
2. We have $\mathbf{x}[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
3. We have $\mathbf{x}[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.

@@

It is worth highlighting that @prp-mtblty assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

## Example: influence of counterfactual training on explainability

@@

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.

We have simulated this scenario using synthetic data with two independent features and Gaussian class-conditional densities in @fig-poc. The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier). In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate counterfactuals in the target class generated at test time using generic gradient descent until convergence.

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with counterfactual training, once again without any mutability constraints. We observe that the counterfactuals are highly plausible, meeting the first objective of @def-explainability.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected the immutable feature as described in @sec-constraints. We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.

@@

![Illustration for @exm-grad of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}
